Modelling BCG vaccination in the UK

### 2019-08-06 rmds/06-tb-bcg-improv-outcomes.Rmd
edits through to chapter 7

To conduct this study, I first obtained all TB notifications for 2009-2015 in England from the Enhanced Tuberculosis surveillance (ETS) system (see Chapter \@ref(data)). I then considered five outcomes: All-cause mortality, death due to TB (in those who died), recurrent TB, pulmonary disease, and sputum smear status. I used logistic regression, with complete case analysis, to investigate each outcome with BCG vaccination, years since vaccination and age at vaccination, adjusting for potential confounders. All analyses were repeated using multiply imputed data. This work was adapted from [@Abbott:2019ir] (also available as a preprint^[Preprint: https://doi.org/10.1101/366476]) supervised by Hannah Christensen and Ellen Brooks-Pollock. Collaborators at Public Health England including, Maeve K Lalor, Dominik Zenner, Colin Campbell, and Mary E Ramsay provided the data and commented on multiple versions of this paper.

@@

In the univariable analysis the odds of death from any cause were lower for BCG vaccinated TB cases compared to unvaccinated cases, with an OR of `r OR.mort` (Table \@ref(tab:06-bcg-mortality), Table \@ref(tab:06-bcg-sum-tab))); an association remained after adjusting for confounders, but was attenuated with an aOR of `r aOR.mort`.  I estimate that if all unvaccinated cases had been vaccinated there would have been on average 19 (95% CI 9 to 29)  fewer deaths per year during the study period (out of 81 deaths per year on average in unvaccinated cases). Whilst there was evidence in univariable analyses to suggest all-cause mortality was higher in persons vaccinated more than 10 years prior to notification of TB and that all-cause mortality increased with increasing age group, these disappeared after adjusting for potential confounders (Table \@ref(tab:06-yrssincebcg-sum-tab), Table \@ref(tab:06-ageatvac-sum-tab)).

@@

```{r 06-imp-yrsincebcg}
sum_imp_yrsinceBCG %>% 
  set_names( c('Outcome', 'OR (95% CI)', 'P-value', 'fmi', 'aOR (95% CI)', 'P-value', 'fmi')) %>% 
  imp_tab_gen(
    caption.short = "ummary of associations between years since vaccination and all outcomes, using pooled imputed data.", 
    caption = 'Summary of associations between years since vaccination and all outcomes, using pooled imputed data. There was insufficient data to estimate an effect for deaths due to TB (in those who died)') %>% 
  add_header_above(c(" " = 1, "Univariable" = 3, "Multivariable" = 3)) %>% 
  kable_styling(latex_options = "HOLD_position")
```

@@

sum_imp_ageatvac %>% 
  set_names( c('Outcome', 'Age group', 'OR (95\\% CI)', 'P-value', 'fmi', 'aOR (95\\% CI)', 'P-value', 'fmi')) %>% 
  imp_tab_gen(caption = 'Summary of associations between age at vaccination and all outcomes, using pooled imputed data (reference is vaccination at <1 year).', escape = FALSE) %>% 
  add_header_above(c(" " = 2, "Univariable" = 3, "Multivariable" = 3))  %>% 
  kable_styling(font_size = 8, latex_options = "HOLD_position") 
```

@@

```{r 06-sens-table, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
bcg_ass_main %>% 
  set_names(c('Study population', 'Outcome', 'BCG', 'OR (95% CI)', 'P-value', 'aOR (95% CI)', 'P-value')) %>% 
  kable(
    caption.short = "Summary of associations between BCG vaccination and all outcomes.", 
    caption = "Summary of associations between BCG vaccination and all outcomes; cases that have no recurrent flag in the ETS (n=50407), and cases that would have been eligible for the BCG schools scheme (n=9943). Those defined to be eligible for the schools scheme are the UK born, that were aged 14 or over in 2004", booktabs = TRUE) %>% 
    column_spec(c(1:2), width = "3cm") %>% 
    kable_styling(latex_options = "HOLD_position") %>%
    add_footnote(label = c("OR: odds ratio with 95% confidence intervals", "aOR: adjusted odds ratio with 95% confidence intervals",
                         "fmi: fraction of missing information",
                         "* Death due to TB in those who died and where cause of death was known"),
               notation = "none") %>% 
    add_header_above(c(" " = 3, "Univariable" = 2, "Multivariable" = 2)) %>% 
    kable_styling(font_size = 8, latex_options = "HOLD_position")
```

@@

Several other Chapters (Chapter \@ref(sutherland), Chapter \@ref(direct-eff), and Chapter \@ref(model-results)) in this thesis assess the impact of moving from universal school age vaccination to selective high risk neonatal vaccination. The reduction in BCG coverage that this implies means that on top of any potential increase in TB incidence rates there may also have been a reduction in the benefical effects from the BCG vaccine discussed in this Chapter. However, as outlined in the previous paragraph, the evidence of reductions in both all-cause, and TB specific mortality, is strongest in the early years of life. This means that the move to neonatal vaccination may have actually led to an increase in the non-specific benefits.

>>>>

To conduct this study, I first obtained all TB notifications for 2009-2015 in England from the Enhanced Tuberculosis surveillance (ETS) system (see Chapter \@ref(data)). I then considered five outcomes: All-cause mortality, death due to TB (in those who died), recurrent TB, pulmonary disease, and sputum smear status. I used logistic regression, with complete case analysis, to investigate each outcome with BCG vaccination, years since vaccination and age at vaccination, adjusting for potential confounders. All analyses were repeated using multiply imputed data. This work was adapted from [@Abbott:2019ir]^[Paper: https://doi.org/10.1016/j.vaccine.2019.06.056] (also available as a preprint^[Preprint: https://doi.org/10.1101/366476]) supervised by Hannah Christensen and Ellen Brooks-Pollock. Collaborators at Public Health England including, Maeve K Lalor, Dominik Zenner, Colin Campbell, and Mary E Ramsay provided the data and commented on multiple versions of this paper.

@@

In the univariable analysis the odds of death from any cause were lower for BCG vaccinated TB cases compared to unvaccinated cases, with an OR of `r OR.mort` (Table \@ref(tab:06-bcg-mortality), Table \@ref(tab:06-bcg-sum-tab)); an association remained after adjusting for confounders, but was attenuated with an aOR of `r aOR.mort`.  I estimate that if all unvaccinated cases had been vaccinated there would have been on average 19 (95% CI 9 to 29)  fewer deaths per year during the study period (out of 81 deaths per year on average in unvaccinated cases). Whilst there was evidence in univariable analyses to suggest all-cause mortality was higher in persons vaccinated more than 10 years prior to notification of TB and that all-cause mortality increased with increasing age group, these disappeared after adjusting for potential confounders (Table \@ref(tab:06-yrssincebcg-sum-tab), Table \@ref(tab:06-ageatvac-sum-tab)).

@@

```{r 06-imp-yrsincebcg}
sum_imp_yrsinceBCG %>% 
  set_names( c('Outcome', 'OR (95% CI)', 'P-value', 'fmi', 'aOR (95% CI)', 'P-value', 'fmi')) %>% 
  imp_tab_gen(
    caption.short = "ummary of associations between years since vaccination and all outcomes, using pooled imputed data.", 
    caption = 'Summary of associations between years since vaccination and all outcomes, using pooled imputed data. There was insufficient data to estimate an effect for deaths due to TB (in those who died)') %>% 
  add_header_above(c(" " = 1, "Univariable" = 3, "Multivariable" = 3))
```

@@

sum_imp_ageatvac %>% 
  set_names( c('Outcome', 'Age group', 'OR (95\\% CI)', 'P-value', 'fmi', 'aOR (95\\% CI)', 'P-value', 'fmi')) %>% 
  imp_tab_gen(caption = 'Summary of associations between age at vaccination and all outcomes, using pooled imputed data (reference is vaccination at <1 year).', escape = FALSE) %>% 
  add_header_above(c(" " = 2, "Univariable" = 3, "Multivariable" = 3))
```

@@

```{r 06-sens-table, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
bcg_ass_main %>% 
  set_names(c('Study population', 'Outcome', 'BCG', 'OR (95% CI)', 'P-value', 'aOR (95% CI)', 'P-value')) %>% 
  kable(
    caption.short = "Summary of associations between BCG vaccination and all outcomes.", 
    caption = "Summary of associations between BCG vaccination and all outcomes; cases that have no recurrent flag in the ETS (n=50407), and cases that would have been eligible for the BCG schools scheme (n=9943). Those defined to be eligible for the schools scheme are the UK born, that were aged 14 or over in 2004", booktabs = TRUE) %>% 
    column_spec(c(1:2), width = "3cm") %>% 
    kable_styling(latex_options = "HOLD_position") %>%
    add_footnote(label = c("OR: odds ratio with 95% confidence intervals", "aOR: adjusted odds ratio with 95% confidence intervals",
                         "fmi: fraction of missing information",
                         "* Death due to TB in those who died and where cause of death was known"),
               notation = "none") %>% 
    add_header_above(c(" " = 3, "Univariable" = 2, "Multivariable" = 2)) 
```

@@

Several Chapters (Chapter \@ref(sutherland), Chapter \@ref(direct-eff), and Chapter \@ref(model-results)) in this thesis assess the impact of moving from universal school age vaccination to selective high risk neonatal vaccination. The reduction in BCG coverage that this implies means that on top of any potential increase in TB incidence rates there may also have been a reduction in the benefical effects from the BCG vaccine discussed in this Chapter. However, as outlined in the previous paragraph, the evidence of reductions in both all-cause, and TB specific mortality, is strongest in the early years of life. This means that the move to neonatal vaccination may have led to an increase in the non-specific benefits.
